---
title: "Project2"
author: "jeromej2"
date: "7/29/2018"
output:
  html_document: default
  pdf_document: default
---
```{r, echo=FALSE, warning=FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")
library(lmtest)
library(knitr)
library(nortest)
library(ggplot2)
library(readr)
library(leaps)
op = function(x, d=4) sprintf(paste0("%1.",d,"f"), x) 
diagnostics <- function(model, 
                        pcol = "grey", 
                        lcol = "dodgerblue", 
                        alpha = 0.05, 
                        plotit = TRUE, 
                        testit = TRUE){
  if (plotit){
    par(mfrow = c(1,2), oma=c(2,0,2,0))
    plot(fitted(model), resid(model), col = pcol, pch = 20,
         xlab = paste("Fitted Values, BP: ", op(bptest(model)$statistic), sep = " "),
         ylab = "Residual Values", 
         main = paste("Fitted vs. Residuals for\n", deparse(substitute(model)), sep = ""))
    abline(h = 0, lwd = 2, col = lcol)
    
    qqnorm(resid(model), col = pcol, pch = 20, main = "Q-Q plot")
    qqline(resid(model), lty = 2, lwd = 2, col = lcol)
  }
  if (testit){
    retframe <- list(p_val = shapiro.test(resid(model))$p.val, 
                           decision = ifelse(shapiro.test(resid(model))$p.val > alpha, 
                                             "Fail to Reject", 
                                             "Reject"))
    retframe
  }
}

```



```{r setup, echo=FALSE, warning=FALSE}

planets <- read_csv("planets_lite_ns.csv")
cols_used <- c("pl_discmethod", "pl_disc", "pl_locale", "st_dist", "st_optmag", "st_teff", "st_mass", 
                 "st_rad", "st_logg", "st_metfe", "st_vsini", "pl_orbsmax", "pl_orbeccen", "pl_bmasse")

planets_used <- planets[,cols_used]
planets_used_graph <- planets_used

factor_cols <- c("pl_discmethod", "pl_disc", "pl_locale")
planets_used[factor_cols] <- lapply(planets_used[factor_cols], factor)

planets_good_idx = complete.cases(planets_used)
planets_good = planets_used[planets_good_idx,]

planets_good <- planets_good[planets_good$pl_disc != 1989,]

op = function(x, d=4) sprintf(paste0("%1.",d,"f"), x) 
diagnostics <- function(model, 
                        pcol = "grey", 
                        lcol = "dodgerblue", 
                        alpha = 0.05, 
                        plotit = TRUE, 
                        testit = TRUE){
  if (plotit){
    par(mfrow = c(1,2), oma=c(2,0,2,0))
    plot(fitted(model), resid(model), col = pcol, pch = 20,
         xlab = paste("Fitted Values, BP: ", op(bptest(model)$statistic), sep = " "),
         ylab = "Residual Values", 
         main = paste("Fitted vs. Residuals for\n", deparse(substitute(model)), sep = ""))
    abline(h = 0, lwd = 2, col = lcol)
    
    qqnorm(resid(model), col = pcol, pch = 20, main = "Q-Q plot")
    qqline(resid(model), lty = 2, lwd = 2, col = lcol)
  }
  if (testit){
    retframe <- list(p_val = shapiro.test(resid(model))$p.val, 
                           decision = ifelse(shapiro.test(resid(model))$p.val > alpha, 
                                             "Fail to Reject", 
                                             "Reject"))
    retframe
  }
}

calc_loocv_rmse = function(model) {
sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

```

## Model selection
I wanted to see if the advances in exoplanetary discovery were getting more accurate in regards to predicting the orbit of extra-solar bodies. Are we getting better in finding panets in the *'Goldilocks Zone'*? Could we find a second home?

The model selection started with a full base fit of the Orbital semi-Major axis as the response and the selected parameters as predictors. Looking at the pairs plot, we can see a few obviously logarithmic relationships between predictors, but for completeness, I started with a simple additive model. I took this model and ran both AIC and BIC to see if we could improve the $R^{2}$. The base model run under backwards BIC returned the best additive model, but the $R^{2}$ was less than 30%.

Next I tried a log transformation on the response, and three of the predictors, I used $e^{predictor}$ of the predictors already in log space, and one interactive term of $log(Stellar Radius * e^{Stellar Surface Gravity})$, as these terms showed high Collinearity and strong Variance Inflation on the base model. The log fit showed promise, so I used both backwards and forwards AIC to see if a better model could be had. In the end of those tests, the original log fit worked out the best.

To verify the fit adhered to normal variance and homoscedacity in the residuals, I ran the shapiro test, and failed to reject the Null hypothesis. The bp test was close, so I plotted a histogram of the residuals and ovelayed a normal curve over for illustration. Although it wasn't perfectly normal, it was very close. I then ran a LOOCV test on the model and got a cross validated root mean squared error of *1.099*, as opposed to base BIC model of *1.268*. There were no outstanding variation inflation factors on this model, and the model itself gave us an adjusted $R^{2}$ of *60.0%*, with a very low p-value.

Lastly, I created a 90% confidence interval for predictions grouped by year. I dropped the one observation from 1989 as I didn't want only one point to overly influence the predictions. I then created a mean confidence interval per year, and plotted the time series. Although not flawless, there appears to be a downward trend in confidence span, inferring that we may be getting more proficient in the discovery of new planets on other systems.


## Methods

```{r}
pairs(planets_good)

```

```{r ModelSelection, echo = FALSE }

base <- lm(pl_orbsmax ~ ., data = planets_good)

base_aic <- step(base, data = planets_good, trace = FALSE)

base_bic <- step(base, data = planets_good, k = log(length(base$residuals)),  trace = FALSE)

log_base <- lm(log(pl_orbsmax) ~ pl_discmethod  + log(st_dist) + log(st_teff) + 
                 log(I(st_rad * exp(st_logg))) + exp(st_metfe) + 
                 st_vsini + exp(pl_orbeccen) + log(pl_bmasse), 
               data = planets_good)


log_base_aic_b <- step(log_base, data = planets_good, trace = FALSE)

start <- lm(log(pl_orbsmax) ~ 1, data = planets_good)
log_base_aic_f <- step(start, 
                       scope = log(pl_orbsmax) ~ pl_discmethod  + log(st_dist) + log(st_teff) +
                         log(I(st_rad * exp(st_logg))) + exp(st_metfe) + 
                         st_vsini + exp(pl_orbeccen) + log(pl_bmasse),
                       direction = 'forward', trace = FALSE)

```

```{r Testing Output}

anova(base,base_aic)
anova(base,base_bic)
summary(base_bic)
calc_loocv_rmse(base_bic)
car::vif(base)
anova(log_base, log_base_aic_b)
anova(log_base, log_base_aic_f)

car::vif(log_base)
summary(log_base)
calc_loocv_rmse(log_base)

best_fit <- log_base

```

## Final Plots

```{r plots}
ggplot(planets_good) + 
  geom_point(aes(x=pl_orbsmax, y=exp(best_fit$fitted.values), color = pl_disc)) + 
  geom_abline(slope = 1,
              intercept = 0,
              col = 'steelblue') + 
  xlab("Real Values") + 
  ylab("Predicted Values") + 
  ggtitle("Predicted vs. Actual Orbital Semi-Major Axis")

diagnostics(best_fit, pcol = 'steelblue', lcol = 'darkorange')


ggplot() + aes(resid(best_fit)) + 
  geom_histogram(aes(y = ..density..), 
                 binwidth=.35, 
                 colour="steelblue", 
                 fill="dark orange") +
  ggtitle("Histogram of Best Fit Residuals\nWith Normal Curve") +
  xlab("Residuals") + 
  ylab("Density") + 
  stat_function(fun=dnorm,
                         color="black",
                         args=list(mean=mean(resid(best_fit)), 
                                   sd=sd(resid(best_fit))))

```

## Confidence intervals By Year (pl_disc)

```{r}

years <- sort(unique(planets_good$pl_disc))
#as.numeric(years)
yearly_predict_range <- data.frame(year = rep(0,length(years)), 
                             span = rep(0,length(years)))

for (i in seq(1:length(years))){
  temp_pred <- exp(predict(best_fit, 
                           newdata = planets_good[planets_good$pl_disc == as.character(years[i]),], 
                           interval = "confidence", 
                           level = 0.90))
  yearly_predict_range[i,] <- c(as.character(years[i]), 
                          mean(temp_pred[,3]) - mean(temp_pred[,2]))
}


ggplot(yearly_predict_range, aes(x=as.Date(year, "%Y"), y=as.numeric(span))) + 
  geom_point(col = 'steelblue') + 
  xlab("Discovery Year") + 
  ylab("Mean 90% Confidence Interval") + 
  ggtitle("Discovery Efficiency By Year") + 
  theme(axis.text.x = element_text(angle=45)) + 
  geom_smooth(method = "lm", se = FALSE,
              col = 'darkorange')

```

